---
title: "WrittenReport"
author: "html_document"
date: "2025-10-08"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, mesage = FALSE)
```

```{r, echo = FALSE}
library(dplyr)
library(cfbfastR)
```

# Executive Summary

Rushing offense is one of the most heavily analyzed aspects of football. The role and impact of a run game in an offense has changed significantly over the past decade. While its role in the NFL is becoming more well defined, college football rushing attacks are utilized in different ways, and one that has yet to be thoroughly evaluated within a quantitative lense. To do this, we aim to quantify Quarterback (QB) and Running Back (RB) impact through Expected Points Added (EPA) to understand individual QB and RB impacts to their teams expected points, and furthermore, their teams wins. We started by gathering play-by-play rushing data and contextual information on all FBS RB's and designed runs/scrambles from QB's from the 2024 regular season. Using past information on RB's and QB's as well as their EPA through the regular season, we aim to predict how much expected points they would add in the future, accounting for the conference they play in, defenses they faced, and other situational factors such as redzone plays, opposing rushing defense latent ability, offensive pass strength and home advantage. 

From our framework, we were able to estimate how many points a QB and RB add over an "average" player, a replacement player. To calculate the value of a replacement level RB, we found the average number of RB's to appear in a game across each conference, and multiplied this value across the number of teams in each conference. Using this product as a threshold, we established that all players above this threshold in terms of snap count were considered non-replacement players, and all players below the threshold were classified as replacement players. Averaging our point estimates for all replacement players in each conference, we were able to establish how many points a replacement RB would add. For QB's, we considered the number of teams as the snap count threshold for replacement vs non replacement (i.e Big Ten has 18 teams, so the top 18 QB's in the Big Ten by snap count were considered non-replacement, and all other QB's were considered as replacement level players).  We then translated points to wins by estimating how much points and score differential have an impact on wins. Using both points added above replacement players by conference (IPAR), as well as wins above replacement players (WAR), we can analyze how much value QB's and RB's add through their rushing abilities. 


# Technical Summary 

We obtained data from the cfbfastR R package and the CollegeFootballData API:
 
 - Play-by-play data (load_cfb_pbp) provided play-level information, including EPA (Expected Points Added), rushing/passing flags, and game context.
 
 - Roster data (load_cfb_rosters) was used to attach player names, teams, and positions.
Team and conference info (cfbd_team_info, cfbd_conferences) identified FBS teams and their conferences.

- Schedules (load_cfb_schedules) provided final scores for converting points to wins.
We filtered to regular season rushes only, then merged rosters to plays:

```{r}
rosters <- load_cfb_rosters(seasons = 2024)
conferences <- cfbd_conferences()
teams <- cfbd_team_info()



pbp <- load_cfb_pbp(seasons = 2024)

teams$conference[teams$school == "Notre Dame"] = "Big Ten"
teams$conference[teams$school == "UConn"] = "ACC"


rushes <- pbp %>%
  filter(rush == 1) %>%
  filter(season_type == "regular") 
rushes$rush_player_id <- as.character(rushes$rush_player_id)
```

Several contextual variables were used to control for game situation and team quality:

- QB run indicator (qbRun) – distinguishes QB scrambles/design runs from RB carries.

- Home field (homeTeam) – 1 if rushing team is at home.

- Pass strength – team’s average EPA per pass, representing balance in offensive scheme.

- Defensive stuff rate – proportion of rushes stopped at or behind the line.

```{r}

```

We estimated the expected contribution of each player to rushing EPA using linear mixed-effects models (lmer). Separate models were fit for QB runs and non-QB runs.

**EPAplay ∼ β0 + β1 ⋅ homeTeam + β2 ⋅ passStrength + β3 ⋅ redZone + β4 ⋅ defStuffRate + uplayer + udefTeam + uconference + ε where uplayer, udefTeam, and uconference are random intercepts.**

```{r}

```

We extracted player random effects and scaled them by usage:

- IPA: contribution per play.

- IPAA: total contribution = IPA × number of plays.

- IPAR: contribution above replacement, defined below.

```{r}

```

We defined replacement level within each conference:

- RBs: identified a threshold based on the average number of RBs per team × number of teams in the conference. RBs below that usage threshold were considered replacement-level.

- QBs: set the replacement threshold equal to the number of teams in the conference.

The average IPA of replacement players in each conference became the replacement baseline. IPAR (Impact above Replacement) is then:

**IPAR = IPAA − (nplays ⋅ avg ⋅ replacement ⋅ IPA)**

To translate points into wins, we regressed team wins on scoring differential:

```{r}

```


**To support our method of using conference as a random effect, we visualized relative conference performance.**

```{r, echo=FALSE}
library(cfbfastR)
library(dplyr)
library(ggplot2)
library(ggimage)
library(tidyr)

logos = data.frame(
  team = c("ACC","American Athletic", "Big 12","Big Ten", "Conference USA", "FBS Independents", "Mid-American", "Mountain West", "Pac-12", "SEC", "Sun Belt"),
  logo = c("https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/1.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/151.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/4.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/5.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/12.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/18.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/15.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/17.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/9.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/8.png&transparent=true&w=30&h=30",
           "https://a.espncdn.com/combiner/i?img=/i/teamlogos/ncaa_conf/500/37.png&transparent=true&w=30&h=30")
)
games = load_cfb_schedules(2024) %>% 
  filter(season_type == "regular", 
         home_division == 'fbs', 
         away_division == 'fbs',
         completed == TRUE,
         (home_points != 0 | away_points != 0),
         home_conference != away_conference)

conf = games %>% 
  select(season,home_team,home_conference) %>%
  rename(team = home_team, conf = home_conference) %>% 
  distinct(conf)

games = games %>% 
  mutate(result = home_points-away_points,
         win_t = ifelse(result > 0, home_conference, away_conference),
         lose_t = ifelse(result < 0, home_conference, away_conference),
         win_by = ifelse(win_t == home_conference, result, -1*result),
         lose_by = ifelse(lose_t == home_conference, result, -1*result) ) %>% 
  select(season,win_t,win_by,lose_t,lose_by,result)

win_diff =
  games  %>% 
  dplyr::group_by(season, win_t) %>%
  dplyr::summarise(wins = dplyr::n(), win_diff = sum(win_by), .groups = 'drop') %>%
  dplyr::rename(team = win_t)

loss_diff =
  games %>%
  dplyr::group_by(season, lose_t) %>%
  dplyr::summarise(loss = dplyr::n(), loss_diff = sum(lose_by), .groups = 'drop') %>%
  dplyr::rename(team = lose_t)

records =
  win_diff %>%
  dplyr::full_join(y = loss_diff) %>%
  dplyr::mutate(scoring_diff = win_diff + loss_diff,
                record = wins-loss) %>% 
  left_join(logos)


ggplot(records,aes(scoring_diff,wins,image = logo))+
  geom_image()+
  labs(x = "Total scoring margin",
       y = "Wins",
       title = "Wins vs Score Differential")+
  theme_bw()

ggplot(records,aes(scoring_diff,loss,image = logo))+
  geom_image()+
  labs(x = "Total scoring margin",
       y = "Loses",
       title = "Loses vs Score Differential")+
  theme_bw()

ggplot(records,aes(scoring_diff,record,image = logo))+
  geom_image(size=.1)+
  labs(x = "Total scoring margin",
       y = "Record",
       title = "Record vs Score Differential")+
  theme_bw()


##########################


games = load_cfb_schedules(2024) %>% 
  filter(season_type == "regular", 
         home_division == 'fbs', 
         away_division == 'fbs',
         completed == TRUE,
         (home_points != 0 | away_points != 0),
         home_conference != away_conference)

conf_games = games %>%
  mutate(point_diff = home_points - away_points) %>% 
  mutate(
    conf_pair = ifelse(home_conference < away_conference,
                       paste(home_conference, away_conference, sep = "_"),
                       paste(away_conference, home_conference, sep = "_")),
    # Flip sign if we swapped the order so that point_diff always corresponds to the first conf in the pair
    point_diff_aligned = ifelse(home_conference < away_conference, point_diff, -point_diff),
    conf1 = ifelse(home_conference < away_conference, home_conference, away_conference),
    conf2 = ifelse(home_conference < away_conference, away_conference, home_conference)
  ) %>% 
  select(conf1,conf2,point_diff_aligned)

avg_diff <- conf_games %>%
  group_by(conf1, conf2) %>%
  summarize(avg_point_diff = mean(point_diff_aligned), .groups = "drop")




logos = rename(logos, conference = team)


plot2_data_a = avg_diff %>% 
  rename(conf_1 = conf2,
         conf_2 = conf1) %>% 
  rename(conf1 = conf_1,
         conf2 = conf_2) %>% 
  mutate(avg_point_diff = -avg_point_diff)

plot2_data = avg_diff %>% 
  full_join(plot2_data_a) %>% 
  left_join(logos, by=c("conf2"="conference")) %>% 
  mutate(avg_point_diff = -avg_point_diff)

ggplot(plot2_data, aes(x=avg_point_diff,y=conf1,image=logo))+
  geom_image(size = 0.04)+
  theme_bw()+
  labs(
    title = "Average Point Differences Between FBS Conferences",
    subtitle = "When the conferences play each other",
    x = "Relative Average Point Differential",
    y = "Conference"
  )

#######

rosters <- load_cfb_rosters(seasons = 2024)
conferences <- cfbd_conferences()
teams <- cfbd_team_info()

pbp <- load_cfb_pbp(seasons = 2024)

rushes <- pbp %>%
  filter(rush == 1) %>%
  filter(season_type == "regular")
rushes$rush_player_id <- as.character(rushes$rush_player_id)


rosterids <- rosters %>%
  mutate(
    name = paste(first_name, last_name, sep = " ")
  ) %>%
  select(name, team, position, athlete_id)


rushes <- rushes %>%
  left_join(rosterids, by = c("rush_player_id" = "athlete_id"))

teaminfo <- rosterids %>%
  left_join(teams, by = c("team" = "school"))

#Replacement player calculation
rbs <- rushes %>%
  filter(position %in% c("RB")) %>%
  group_by(pos_team, position, conference, rusher_player_name) %>%
  summarize(
    rush_attempts = n(),
    total_rush_epa = sum(EPA, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  group_by(conference) %>% 
  summarize(
    avg_epa_conf = mean(total_rush_epa, na.rm = TRUE, .groups = "drop")
  ) %>% 
  filter(conference %in% c("American Athletic", "ACC", "Big 12", "Big Ten", "Conference USA",
                           "FBS Independents", "Mid-American", "Mountain West", "Pac-12",
                           "SEC", "Sun Belt")) %>% 
  left_join(logos)

ggplot(rbs)+
  geom_image(aes(x = conference, y=avg_epa_conf, image = logo), position = position_jitter(height = .2))+
  geom_point(aes(y=-.5,x=0), alpha = 0)+
  geom_point(aes(y=.5,x=0), alpha = 0)+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  labs(
    title = "Average EPA Per Conference",
    x = "Conference",
    y = "Average Player EPA"
  )
```


Lastly, we computed WAR as:

**WAR = IPAR ⋅ pointsToWin**

```{r}

```

Here were our results:

```{r}

```

```{r}

```

Limitations:

- This WAR metric only accounts for rushing, not passing or receiving due to lack of play-by-play data provided at the college level. 

- Since there are only 2 independent teams, Notre Dame was moved to Big Ten and UConn was moved to ACC. 

- Replacement thresholds are heuristics; applying different definitions of replacement could change the results.

- The linear mixed-effects model does not capture interactive variables such as QB mobility vs defensive schemes.

- Only the regular season statistics were used during this development.

Next Steps:


